The Ranked Nearest Neighbour Interchange space

### 2019-06-26 paper.tex
define splits + minor

A \emph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \to \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$ or $rank_T(v)$ if it is not obvious that we consider tree $T$.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
Each internal node $v$ gives a \emph{cluster} $C$ that is the set of taxa descending from this node. \todo{OK?}
We then say that $v$ induces the cluster $C$.
Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

\todo{define splits and delete definition from split thm chapter!}

@@

%Is it sufficient to mention what RNNI stands for in the Introduction?

%Restrict to RNNIu (See whether we can get same results for RNNI)

%Definitions inside a tree: cherry, partition, edges represent partitions

@@

%delete definitions in the following chapters that are already covered here!

%replace rank swap with rank move in the following!!

>>>>

A \emph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \to \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$ or $rank_T(v)$ if it is not obvious that we consider tree $T$.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.

Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
Each internal node $v$ gives a \emph{cluster} $C$ that is the set of taxa descending from this node. \todo{OK?}
We then say that $v$ induces the cluster $C$.
Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
A \emph{split} is a bipartition of the taxon set $X$ into sets $A,B$ such that there is an edge $e$ whose deletion results in two trees with taxon sets $A$ and $B$, respectively.
Such a split will be denoted by $A|B$ and we say that it is induced by edge $e$.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes

@@

@@
